A hazards approach to the biometric analysis of infant mortality

### 2023-03-08 paper2.Rmd
cleaning and adding reference

---
title: "A hazards approach to the biometric analysis of infant mortality"
author: "Göran Broström and Tommy Bengtsson"
date: "`r Sys.time()`"
output: 
  bookdown::pdf_document2:
    citation_package: natbib
    keep_tex: yes
    number_sections: yes
    toc: yes
    toc_depth: 3
  bookdown::html_document2:
    number_sections: yes
    toc: yes
    toc_depth: 2
    toc_float: yes
  bookdown::word_document2:
    toc: no
    toc_depth: 2
citation_package: natbib
##classoption: titlepage
bibliography: bb.bib
titlepage: no
biblio-style: apalike
documentclass: article
fontsize: 11pt
header-includes:
- \usepackage{a4wide}
- \usepackage[utf8]{inputenc}
- \usepackage{graphicx}
- \usepackage{sectsty}
- \usepackage[figuresonly]{endfloat}
- \renewcommand{\efloatseparator}{\mbox{}}
##- \subsectionfont{\normalfont\normalfont\itshape}
abstract: "A variation of the Bourgeois-Pichat biometric analysis of infant mortality is suggested. In the original model, cumulative mortality in the last eleven months of infancy is assumed to follow a uniform distribution given a log-cube transformation of age. Instead, we assume an exponential distribution. The difference is that while the denominator is constant in the Bourgeois-Pichat model, equal to the number of births, in our model, the denominator is the current population at risk. We argue that our assumption is more satisfactory from a theoretical point of view, since it focus on the conditional probability of dying. Our model gives different estimates of endogenous and exogenous mortality and, in addition, the model ﬁt is slightly better, especially in cases with higher levels of infant mortality."
---

@@

Applying this method, Bourgeois-Pichat also finds deviations from the linear death pattern after the first month. To take one example, the curve for Sardinia in 1948 bend strongly upward after four months [@bp52, Figure 12]. To take another example, in the case of Quebec 1944--1947, the curve bends down quite strongly after the sixth month [@bp52, Figure 8]. A downward bend is also what has been found using historical Swedish parish data [@tb99, Figure 6b; @tbcl99, Figures 6 and 7; @lyn98, Figures 1 and 2; @jslgt81, Figure 6d], though less pronounced than in Quebec. The curve for Sweden 1910--1946 show, however, no such downward bend [@bp51b, Figure 9; @bp52].

@@

```{r postneo, fig.cap = "Exponential fits to postneonatal data, Västerbotten."}
oldpar <- par(mfrow = c(1, 2), cex = 0.7)
post <- age.window(ginfant, c(g(28), g(365)))
p1 <- post[post$period == "1861-1890", ]
p3 <- post[post$period == "1921-1950", ]
fit1 <- coxreg(Surv(exit - enter, event) ~ 1, data = p1)
wfit1 <- phreg(Surv(exit - enter, event) ~ 1, data = p1, shape = 1)
##check.dist(fit1, wfit1, main = "1861-1890", printLegend = FALSE)
plot(fit1, main = "1861-1890", printLegend = FALSE, xlab = "g(Day)", col = "blue",
     ylab = "Cumulative hazards")
whaz1 <- hazards(wfit1)
with(whaz1, lines(x, y, lty = 2, col = "red"))
legend("topleft", legend = c("Non-parametric", "Exponential"), lty = 1:2, 
       col = c("blue", "red"), cex = 1)
##
fit3 <- coxreg(Surv(exit - enter, event) ~ 1, data = p3)
wfit3 <- phreg(Surv(exit - enter, event) ~ 1, data = p3, shape = 1)
whaz <- hazards(wfit3)
plot(fit3, main = "1921-1950", printLegend = FALSE, xlab = "g(Day)")
with(whaz, lines(x, y, lty = 2, col = "red"))
legend("topleft", legend = c("Non-parametric", "Exponential"), lty = 1:2, 
       col = c("blue", "red"), cex = 1)
par(oldpar)
```

@@

2.   On the full infant interval, estimate the total cumulative hazard rate $A(t)$ with the usual 
Nelson-Aalen estimator $\hat{A}(t)$ [@nelson72; @aalen78]. Then, subtract $H(t)$ from 
$\hat{A}(t)$ to get $\hat{E}(t)$.
\begin{equation}
\hat{E}(t) = \hat{A}(t) - \hat{\lambda} t,\; 0 < t \le g(28).
\end{equation}
Note that subtraction and addition of competing risks are okay on the hazards scale, 
but not with probabilities, which is yet another argument in favor of the hazards approach.

@@

Post-neonatal mortality is the same as exogenous mortality in the sense that the
hazard functions are the same on the post-neonatal age interval. So the first step
involves data left truncated at age 28 days. The maximum-likelihood estimator of the rate 
in the exponential distribution is simply the number of observed post-neonatal deaths
`r days`, 
divided by the total post-neonatal exposure time `r expo` (still on the $g$ scale), 
giving a rate of `r rate`.

@@

Here we focus on neonatal period.

The estimation of the cumulative hazard function for the total neonatal mortality is
easily achieved by calling the function `hazards` in the *eha* package, see also Figure \@ref(fig:hazards).

```{r hazards, fig.cap = "Cumulative hazard functions for neonatal mortality.", echo = TRUE}
neo <- age.window(ginfant, c(0, g(28)))
par(las = 1)
fit <- coxreg(Surv(enter, exit, event) ~ 1, data = neo)
hneo <- hazards(fit, cum = TRUE)
x <- c(0, hneo[[1]][, 1])
y <- c(0, 1000 * cumsum(hneo[[1]][, 2]))
plot(x, y, type = "l", xlab = "Day", ylab = "by 1000")
lines(x, 1000 * rate * x, col = "blue", lty = 2)
abline(h = 0, v = 0)
text(8, 15, "Total", col = "black")
text(8, 5, "Exogenous", col = "blue")
```

Next, take the difference between "Total" and "Exogenous" to get "Endogenous", Figure \@ref(fig:enof).

```{r enof, fig.cap = "Cumulative hazard function for endogenous neonatal mortality.", fig.width = 7}
yen <- y - 1000 * rate * x
par(las = 1)
plot(x, yen, type = "l", col = "red", xlab = "Day", ylab = "by 1000", ylim = c(0, 36))
abline(h = 0, v = 0)
lines(x, 1000 * rate * g(x), col = "blue", lty = 3)
lines(x, y, lty = 1)
text(8, 20, "Total", cex = 0.8)
text(15, 14, "Endogenous", col = "red")
text(8, 5, "Exogenous", col = "blue", cex = 0.8)
n <- length(yen)
axis(4, at = c(0, 1000 * rate * g(x[n]), yen[n], y[n]), las = 1, cex.axis = 0.7)
```

@@

## Västerbotten 1921--1950

@@

## Västerbotten 1890--1920

@@

## Västerbotten 1861--1890

@@

The crude IMR in Umeå 1861--1890 was around `r crudprob` per thousand, highest of the three
time periods.
The biometric analysis is shown in 
Figure \@ref(fig:bioume6190).

@@

The conclusion here is almost the same as  for the later time periods, good model
fit and endogenous dominance in the very early days of life. However, the B-P method seems to have a slight upper hand.

@@

@@

Very bad fits in both cases.  <!--- Let us look at the estimated cumulative hazards, total, 
exogenous and endgenous, see Figure \@ref(fig:escu).

@@

An example of severe under-registration during the neonatal period, see Figure \@ref(fig:skum).
There is no room for endogeneous deaths at all, but otherwise a reasonably good fit with both 
methods.

>>>>

---
title: "A hazards approach to the biometric analysis of infant mortality"
author: "Göran Broström and Tommy Bengtsson"
date: "`r Sys.time()`"
output: 
  bookdown::pdf_document2:
    citation_package: natbib
    keep_tex: yes
    number_sections: yes
    toc: yes
    toc_depth: 3
  bookdown::html_document2:
    number_sections: yes
    toc: yes
    toc_depth: 2
    toc_float: yes
  bookdown::word_document2:
    toc: no
    toc_depth: 2
citation_package: natbib
##classoption: titlepage
bibliography: bb.bib
titlepage: no
biblio-style: apalike
setcitestyle: authoryear
documentclass: article
fontsize: 11pt
header-includes:
- \usepackage{natbib}
- \setcitestyle{authoryear}
- \usepackage{a4wide}
- \usepackage[utf8]{inputenc}
- \usepackage{graphicx}
- \usepackage{sectsty}
- \usepackage[figuresonly]{endfloat}
- \renewcommand{\efloatseparator}{\mbox{}}
##- \subsectionfont{\normalfont\normalfont\itshape}
abstract: "A variation of the Bourgeois-Pichat biometric analysis of infant mortality is suggested. In the original model, cumulative mortality in the last eleven months of infancy is assumed to follow a uniform distribution given a log-cube transformation of age. Instead, we assume an exponential distribution. The difference is that while the denominator is constant in the Bourgeois-Pichat model, equal to the number of births, in our model, the denominator is the current population at risk. We argue that our assumption is more satisfactory from a theoretical point of view, since it focus on the conditional probability of dying. Our model gives different estimates of endogenous and exogenous mortality and, in addition, the model ﬁt is slightly better, especially in cases with higher levels of infant mortality."
---

@@

Applying this method, Bourgeois-Pichat also finds deviations from the linear death pattern after the first month. To take one example, the curve for Sardinia in 1948 bend strongly upward after four months [@bp52, Figure 12]. To take another example, in the case of Quebec 1944--1947, the curve bends down quite strongly after the sixth month [@bp52, Figure 8]. A downward bend is also what has been found using historical Swedish parish data [@tb99, Figure 6b; @tbcl99, Figures 6 and 7; @lyn98, Figures 1 and 2; @jslgt81, Figure 6d], though less pronounced than in Quebec. @knoki77 argues that breastfeeding patterns have a strong impact on the linearity. The curve for Sweden 1910--1946 show, however, no such downward bend [@bp51b, Figure 9; @bp52].

@@

```{r postneo, fig.cap = "Exponential fits to postneonatal data, Västerbotten."}
oldpar <- par(mfrow = c(1, 2), cex = 0.7)
post <- age.window(ginfant, c(g(28), g(365)))
p1 <- post[post$period == "1861-1890", ]
p3 <- post[post$period == "1921-1950", ]
fit1 <- coxreg(Surv(exit - enter, event) ~ 1, data = p1)
haz1 <- hazards(fit1)
x <- c(0, haz1[[1]][, 1])
y <- c(0, cumsum(haz1[[1]][, 2]))
##       
wfit1 <- phreg(Surv(exit - enter, event) ~ 1, data = p1, shape = 1)
##check.dist(fit1, wfit1, main = "1861-1890", printLegend = FALSE)
plot(x, y, main = "1861-1890", xlab = "Day", col = "blue",
     ylab = "Cumulative hazards", axes = FALSE, type = "s")
axis(2, las = 1)
axis(1, at = c(0, g(365) - g(28)), labels = c(28, 365))
abline(h = 0)
box()
whaz1 <- hazards(wfit1)
with(whaz1, lines(x, y, lty = 2, col = "red"))
legend("topleft", legend = c("Non-parametric", "Exponential"), lty = 1:2, 
       col = c("blue", "red"), cex = 1)
##
fit3 <- coxreg(Surv(exit - enter, event) ~ 1, data = p3)
haz3 <- hazards(fit3)
x <- c(0, haz3[[1]][, 1])
y <- c(0, cumsum(haz3[[1]][, 2]))
wfit3 <- phreg(Surv(exit - enter, event) ~ 1, data = p3, shape = 1)
whaz <- hazards(wfit3)
plot(x, y, main = "1921-1950", xlab = "Day", col = "blue", 
     ylab = "Cumulativa hazards", axes = FALSE, type = "s")
axis(2, las = 1)
axis(1, at = c(0, g(365) - g(28)), labels = c(28, 365))
box()
abline(h = 0)
with(whaz, lines(x, y, lty = 2, col = "red"))
legend("topleft", legend = c("Non-parametric", "Exponential"), lty = 1:2, 
       col = c("blue", "red"), cex = 1)
par(oldpar)
```

@@

2.   On the full infant interval, estimate the total cumulative hazard rate $A(t)$ with the usual 
Nelson-Aalen estimator $\hat{A}(t)$ [@nelson72; @aalen78]. Then, subtract $H(t)$ from 
$\hat{A}(t)$ to get $\hat{E}(t)$.
\begin{equation} (\#eq:endog)
\hat{E}(t) = \hat{A}(t) - \hat{\lambda} t,\; 0 < t \le g(28).
\end{equation}
Note that subtraction and addition of competing risks are okay on the hazards scale, 
but not with probabilities, which is yet another argument in favor of the hazards approach.

@@

Post-neonatal mortality is the same as exogenous mortality in the sense that the
hazard functions are the same on the post-neonatal age interval. So the first step
involves data left truncated at age 28 days. The maximum-likelihood estimator of the rate 
in the exponential distribution is simply the number of observed post-neonatal deaths
`r as.integer(days)`, 
divided by the total post-neonatal exposure time `r format(round(expo), scientific = FALSE)` (still on the $g$ scale), 
giving a rate of `r rate`.

@@

The estimation of the cumulative hazard function for the total neonatal mortality is standard. <!--
easily achieved by calling the function `hazards` in the *eha* package, see also Figure \@ref(fig:hazards).
-->
```{r hazards, eval = TRUE}
neo <- age.window(ginfant, c(0, g(28)))
par(las = 1)
fit <- coxreg(Surv(enter, exit, event) ~ 1, data = neo)
hneo <- hazards(fit, cum = TRUE)
x <- c(0, hneo[[1]][, 1])
y <- c(0, 1000 * cumsum(hneo[[1]][, 2]))
##plot(x, y, type = "s", xlab = "Day", ylab = "by 1000", axes = FALSE)
##axis(2, las = 1)
##axis(1, at = c(g(0), g(7), g(14), g(21), g(28)), labels = c(0, 7, 14, 21, 28))
##axis(2)
##box()
##lines(x, 1000 * rate * x, col = "blue", lty = 2)
##abline(h = 0, v = 0)
##text(8, 15, "Total", col = "black")
##text(8, 5, "Exogenous", col = "blue")
```

Next, take the difference between "Total" and "Exogenous", as in Equation \@ref(eq:endog) to get "Endogenous",
the result is shown in Figure \@ref(fig:enof).

```{r enof, fig.cap = "Cumulative hazard functions for neonatal mortality, Västerbotten.", fig.width = 7}
yen <- y - 1000 * rate * x
par(las = 1)
plot(x, yen, type = "s", col = "red", xlab = "Day", ylab = "by 1000", ylim = c(0, 36), axes = FALSE)
axis(2, las = 1)
axis(1, at = c(g(0), g(7), g(14), g(21), g(28)), labels = c(0, 7, 14, 21, 28))
box()
abline(h = 0, v = 0)
lines(x, 1000 * rate * g(x), col = "blue", lty = 3)
lines(x, y, lty = 1, type = "s")
text(8, 20, "Total", cex = 0.8)
text(15, 14, "Endogenous", col = "red")
text(8, 5, "Exogenous", col = "blue", cex = 0.8)
n <- length(yen)
axis(4, at = c(0, 1000 * rate * g(x[n]), yen[n], y[n]), las = 1, cex.axis = 0.7)
```

@@

## Västerbotten

Thee data set from Västerbotten 1861--1950, kept by CEDAR, Umeå University [@u22003], 
shows almost perfect fits to both the hazards model
and the original biometric model [@bp51a; @bp51b] in all the three subperiods, see
the results in the following three subsections.

### The period 1921--1950

@@

### The period 1890--1920

@@

### The period 1861--1890

@@

The crude IMR in Västerbotten 1861--1890 was around `r crudprob` per thousand, highest of the three
time periods.
The biometric analysis is shown in 
Figure \@ref(fig:bioume6190).

@@

The conclusion here is almost the same as  for the later time periods, good model
fit and endogenous dominance in the very early days of life. 
However, the B-P method seems to have a slight upper hand regarding model fit.

@@

This is an interesting case because it appears as if the data quality is low since 
the linearity during the post-neonatal period is gone. Rather, there is a "hockey-stick" pattern, 
also seen in other old data sets. See Figure \@ref(fig:scahigh).

However, this data set is known to be of very high quality [@belu99], so we are inclined to 
think that the biometric model as it is described here simply is inappropriate during the eighteenth century.

@@

<!--- Let us look at the estimated cumulative hazards, total, 
exogenous and endgenous, see Figure \@ref(fig:escu).

@@

An example of severe under-registration during the neonatal period, see Figure \@ref(fig:skum).
There is no room for endogenous deaths at all, but otherwise a reasonably good fit with both 
methods.
